Reproducible development environments with rix

### 2023-12-22 paper.qmd
intro to nix done?

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, and the entirety of CRAN and Bioconductor is available
through [Nix]{.proglang}’s repositories. This means that using [Nix]{.proglang},
it is possible to install not only [R]{.proglang}, but also all the packages
required for a project. The reason why one should use [Nix]{.proglang} to
install [R]{.proglang} packages, and not use the usual, built-in,
`install.packages()` function instead, is that when installing a package with
[Nix]{.proglang}, [Nix]{.proglang} makes sure to install every dependency of
every package, whether this dependency is another R package or a system-level
dependency.

For example, the [xlsx]{.pkg} package requires the [Java]{.proglang} programming
language to be installed on your computer to successfully install. Depending on
the system one tries to install [xlsx]{.pkg}, installing [Java]{.proglang} might
not easy, nor even possible. But with [Nix]{.proglang}, it suffices to declare
that the [xlsx]{.pkg} package is needed for the project, and [Nix]{.proglang}
figures out automatically that [Java]{.proglang} is required and installs and
configures it. It all just happens without any required intervention from the
user.

But where do these packages come from? When installing a package using Nix an
expression written in the Nix programming language gets downloaded from the
`nixpkgs` Github repository and evaluated. The evaluation of these expressions
results in the installation of the package, alongside its dependencies. Because
`nixpkgs` is a collection of expressions hosted on Github, it is possible to pin
a specific revision of `nixpkgs` to ensure reproducibility of our project.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
installed.

With Nix, it is essentially possible to replace {renv} and Docker combined, or
if you’re using mainly Python, you can replace `conda` or `requirements.txt`
files. If you need other tools or languages like Python or Julia, this can also
be done easily. Nix is available for Linux, macOS and Windows (via WSL2).
Important remark: since using Nix on Windows must go through WSL, when we refer
to "Linux" in the context of Nix, this includes Windows by default as well. It
is also possible to build multi-language environments, containing R and Python,
a LaTeX distribution and packages and so on.

## The Nix programming language

Nix is not just useful because it is possible to install many packages and even
install older packages, but also because it comes with a complete functional
programming language. This programming language is used to write *expressions*,
and these expressions in turn are used to build software. Essentially, when you
install a package using Nix, an expression gets downloaded from the Nix package
repository (more on that in the next section), and it gets evaluated by the Nix
package manager. This expression contains a so-called *derivation*. A derivation
defines a build: some inputs, some commands, and then an output. Most of the
time, a derivation downloads source code, builds the software from the source
and then outputs a compiled binary. Derivations are extremely flexible, and you
could write a derivation to build a complete environment and then build a
complete reproducible pipeline. The output could be any of the discussed data
products.

Learning the Nix programming language is a good idea if you want to contribute
to the Nix package repository, but you might not have to learn it in-depth if
you simply wish to use it to build reproducible environments, as we will learn
now. If you wish to learn about the programming language, I highly recommend [a
tour of
Nix](https://nixcloud.io/tour/?id=introduction/nix)^[https://nixcloud.io/tour/?id=introduction/nix].

## The Nix package repository

So, there’s the Nix package manager, the Nix programming language and the Nix
package repository (henceforth nixpkgs). To look for packages click
[here](https://search.nixos.org/packages)^[https://search.nixos.org/packages].
The source code of all the packages (so the whole set of Nix expressions) can be
found on [this Github
repository](https://github.com/NixOS/nixpkgs)^[https://github.com/NixOS/nixpkgs].
For example,
[here](https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix)^[https://github.com/NixOS/nixpkgs/blob/master/pkgs/development/libraries/quarto/default.nix]
is the Nix expression that contains the derivation to build `quarto`. As you can
see, the derivation uses the the pre-built Quarto binaries instead of building
it from source. Adding packages to nixpkgs (or updating them) can be done by
opening pull requests. For example,
[here](https://github.com/NixOS/nixpkgs/pull/259443)^[https://github.com/NixOS/nixpkgs/pull/259443]
is a pull request to make Quarto available to all platforms (before this PR
Quarto was only available for Linux). PRs get reviewed and approuved by
maintainers that also have the right to merge the PR into master. Once merged,
the new or updated package is available for download. Because nixpkgs is a
“just” Github repository, it is possible to use a specific commit hash to
install the packages as they were at a specific point in time. For example, if
you use this commit, 7c9cc5a6e, you’ll get the packages as of the 19th of
October 2023, but if you used this one instead: 976fa3369, you’ll get packages
from the 19th of August 2023. Using specific hashes is called "pinning" and you
can read more about it [here](https://nixos.wiki/wiki/FAQ/Pinning_Nixpkgs). We
will make extensive use of pinning.

[R]{.proglang} provides a very flexible implementation of the general GLM framework in the function [glm]{.fct} @ChambersHastie1992 in the [stats]{.pkg} package. Its most important arguments are

```r
glm(formula, data, subset, na.action, weights, offset,
  family = gaussian, start = NULL, control = glm.control(…),
  model = TRUE, y = TRUE, x = FALSE, …)
```

where `formula` plus `data` is the now standard way of specifying regression relationships in [R]{.proglang}/[S]{.proglang} introduced in @ChambersHastie1992. The remaining arguments in the first line (`subset`, `na.action`, `weights`, and `offset`) are also standard for setting up formula-based regression models in [R]{.proglang}/[S]{.proglang}. The arguments in the second line control aspects specific to GLMs while the arguments in the last line specify which components are returned in the fitted model object (of class [glm]{.class} which inherits from [lm]{.class}). For further arguments to [glm]{.fct} (including alternative specifications of starting values) see `?glm`. For estimating a Poisson model `family = poisson` has to be specified.

::: callout
As the synopsis above is a code listing that is not meant to be executed, one can use either the dedicated `{Code}` environment or a simple `{verbatim}` environment for this. Again, spaces before and after should be avoided.

Finally, there might be a reference to a `{table}` such as @tbl-overview. Usually, these are placed at the top of the page (`[t!]`), centered (`\centering`), with a caption below the table, column headers and captions in sentence style, and if possible avoiding vertical lines.
:::

| Type           | Distribution | Method   | Description                                                                                                                                                                                  |
|---------|---------|---------|-----------------------------------------------|
| GLM            | Poisson      | ML       | Poisson regression: classical GLM, estimated by maximum likelihood (ML)                                                                                                                      |
|                |              | Quasi    | "Quasi-Poisson regression'': same mean function, estimated by quasi-ML (QML) or equivalently generalized estimating equations (GEE), inference adjustment via estimated dispersion parameter |
|                |              | Adjusted | "Adjusted Poisson regression'': same mean function, estimated by QML/GEE, inference adjustment via sandwich covariances                                                                      |
|                | NB           | ML       | NB regression: extended GLM, estimated by ML including additional shape parameter                                                                                                            |
| Zero-augmented | Poisson      | ML       | Zero-inflated Poisson (ZIP), hurdle Poisson                                                                                                                                                  |
|                | NB           | ML       | Zero-inflated NB (ZINB), hurdle NB                                                                                                                                                           |
: Overview of various count regression models. The table is usually placed at the top of the page (`[t!]`), centered (`centering`), has a caption below the table, column headers and captions are in sentence style, and if possible vertical lines should be avoided. {#tbl-overview}

## Illustrations {#sec-illustrations}

For a simple illustration of basic Poisson and NB count regression the
`quine` data from the [MASS]{.pkg} package is used. This provides the number
of `Days` that children were absent from school in Australia in a
particular year, along with several covariates that can be employed as regressors. The data can be loaded by

```{R}
#| prompt: true
data(mtcars)
```
and a basic frequency distribution of the response variable is displayed in
@fig-quine.

:::{.callout}
For code input and output, the style files provide dedicated environments.
Either the "agnostic" `{CodeInput}` and `{CodeOutput}` can be used
or, equivalently, the environments `{Sinput}` and `{Soutput}` as
produced by [Sweave]{.fct} or [knitr]{.pkg} when using the `render_sweave()`
hook. Please make sure that all code is properly spaced, e.g., using
`y = a + b * x` and _not_ `y=a+b*x`. Moreover, code input should use "the usual" command prompt in the respective software system. For
[R]{.proglang} code, the prompt `R> ` should be used with `+  ` as
the continuation prompt. Generally, comments within the code chunks should be
avoided -- and made in the regular {{< latex >}} text instead. Finally, empty lines before and after code input/output should be avoided (see above).
:::

## Summary and discussion {#sec-summary}

As usual…

## Computational details {.unnumbered}

If necessary or useful, information about certain computational details
such as version numbers, operating systems, or compilers could be included
in an unnumbered section. Also, auxiliary packages (say, for visualizations,
maps, tables, …) that are not cited in the main text can be credited here.

The results in this paper were obtained using [R]{.proglang}~3.4.1 with the
[MASS]{.pkg}~7.3.47 package. [R]{.proglang} itself and all packages used are available from the Comprehensive [R]{.proglang} Archive Network (CRAN) at
[https://CRAN.R-project.org/].

>>>>

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, and the entirety of CRAN and Bioconductor is available
through [Nix]{.proglang}’s repositories. This means that using [Nix]{.proglang},
it is possible to install not only [R]{.proglang}, but also all the packages
required for a project. The reason why one should use [Nix]{.proglang} to
install [R]{.proglang} packages, and not use the usual, built-in,
`install.packages()` [R]{.proglang} function instead, is that when installing a
package with [Nix]{.proglang}, [Nix]{.proglang} makes sure to install every
dependency of every package, whether this dependency is another [R]{.proglang}
package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on the what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not easy, nor even possible. But with [Nix]{.proglang},
it suffices to declare that the [xlsx]{.pkg} package is needed for the project,
and [Nix]{.proglang} figures out automatically that [Java]{.proglang} is
required and installs and configures it. It all just happens without any
required intervention from the user.

But where do these packages come from? When installing a package using
[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
environment for a project in a reproducible manner.

Because the whole set of [Nix]{.proglang} expression is hosted on Github, it is
possible to pin a specific revision of `nixpkgs` to ensure reproducibility of
our project. Pinning a revision ensures that every package that Nix installs
will always be at exactly the same versions, regardless of when in the future
the packages get installed.

With Nix, it is essentially possible to replace [renv]{.pkg} and Docker
combined, or in the case of [Python]{.proglang}, replace `requirements.txt`
files which list package dependencies for [Python]{.proglang} projects. It is
also possible to build multi-language environments, containing R and Python, a
LaTeX distribution, and even install one's favourite text editor to edit the
project's source code.
